---
title: "Assignment0"
format: html
editor: visual
---

## Exercise 1

Given a binomial random variable X, with probability of success θ = 0.6, compute, for 10 trials (this is specified in R with size = 10):

• Prob(X ≤ 6)

```{r}
pbinom(6, size = 10, prob = 0.6)
```

• Prob(X ≥ 4)

```{r}
pbinom(3, size = 10, prob = 0.6, lower.tail = FALSE)

# cross check
1 - pbinom(3, size = 10, prob = 0.6)
```

• Prob(7 ≤ X ≤ 9)

```{r}
sum(dbinom(7:9, size = 10, prob = 0.6))

# cross check
pbinom(9, size = 10, prob = 0.6) - pbinom(6, size = 10, prob = 0.6)
```

## Exercise 2

Given a binomial random variable X, with probability of success θ = 0.5, what is the quantile k such that Prob(X ≤ k) = 0.5. The number of trials is 10.

```{r}
qbinom(p = 0.5, size = 10, prob = 0.5)
```

## Exercise 3

Run an experiment with 10 trials from a binomial (probability of success 0.6) for 1000 times. Save the 1000 instances as an R object (say, successes).

```{r}
successes <- rbinom(n = 1000, size = 10, prob = 0.6)
```

Compute the variance of the successes vector (use the command var()).

```{r}
var(successes)
```

Compute the variance of the number of successes analytically using the formula nθ(1 − θ) shown in the slides. Confirm that the two estimates (the one based on simulation and the one based on the analytical calculation) approximately match.

```{r}
10 * 0.6 * (1 - 0.6)
```

Plot the distribution of the number of successes for the 1000 experiments.

```{r}
hist(successes)
```

## Exercise 4

Given a normal random variable X with mean 500 and variance $50^2$, compute:

• Prob(X \> 550)

```{r}
pnorm(550, mean = 500, sd = 50, lower.tail = FALSE)

# cross check
1 - pnorm(550, mean = 500, sd = 50)
```

• Prob(X \< 450)

```{r}
pnorm(450, mean = 500, sd = 50)
```

• Prob(450 \< X \< 550)

```{r}
pnorm(550, mean = 500, sd = 50) - pnorm(450, mean = 500, sd = 50)
```

• Prob(450 ≤ X ≤ 550)

```{r}
pnorm(550, mean = 500, sd = 50) - pnorm(450, mean = 500, sd = 50)
```

## Exercise 5

Given a normal random variable X with mean 500 and variance $50^2$:

• what is the quantile q such that Prob(X \< q) = 0.80?

```{r}
qnorm(p = 0.8, mean = 500, sd = 50)
```

• what is the quantile q such that Prob(X \> q) = 0.30?

```{r}
# Prob(X > q) = 0.30 is equivalent to Prob(X < q) = 0.70

qnorm(p = 0.7, mean = 500, sd = 50)
```

• what are the two quantiles q1 and q2 such that they are equidistant from the mean 500 and Prob(q1 \< X \< q2) = 0.95?

```{r}
# Prob(q1 < X < q2) = 0.95 is equivalent to Prob(q1 < X) = 0.025 and Prob(q2 > X) = 0.025, with the latter equivalent to Prob(q2 < X) = 0.975

qnorm(p = 0.025, mean = 500, sd = 50)
qnorm(p = 0.975, mean = 500, sd = 50)
```

## Exercise 6

Generate 100 data points from a normal distribution with mean 500 and variance $50^2$ for 1000 times. Each time, compute the mean of the 100 data points and save it in an object.

```{r}
means <- rep(NA, 1000)
for(i in 1:1000){
  means[i] <- mean(rnorm(100, mean=500, sd=50))
}
```

Confirm that the mean of the vector of means is close to 500.

```{r}
mean(means)
```

Confirm that the standard deviation of the vector of means is approximately the analytically computed standard error $\frac{50}{\sqrt{100}}$

```{r}
sd(means)

# analytical SE
50 / sqrt(100)
```

Plot the distribution of the means.

```{r}
hist(means)
```
