---
title: "Assignment4"
format: html
editor: visual
---

```{r setup}
#| message: false
#| warning: false
  
# packages
library(tidyverse) # Graphs with ggplot()
library(knitr) # Tables with kable()
library(brms) # Hierarchical Bayesian models
library(rstan) # Interface to Stan
# library(shinystan) # Model output graphs
library(bayesplot) # Model output graphs

# general setting
rstan_options(auto_write = TRUE)
options(contrasts = c("contr.equalprior_pairs", "contr.poly"))
set.seed(359)
```

The data consist of several measurements of penguins of the Palmer archipelago in Antarctica.

```{r}
#| message: false
#| warning: false

library(palmerpenguins)
data(penguins)

ggplot(penguins, 
             aes(x = bill_length_mm, y = flipper_length_mm, 
                 color = species))+
  geom_point(alpha = .5)+
  scale_color_viridis_d(begin = .1, end = .9)+
  labs(x = "Bill length", y = "Flipper length", color = "Species")+
  theme_bw()

# Remove missing values
penguins <- na.omit(penguins)

# Center numeric predictors in case you build models with interaction terms
penguins$bill_len_c <- 
  penguins$bill_length_mm - mean(penguins$bill_length_mm)
penguins$bill_dep_c <- 
  penguins$bill_depth_mm - mean(penguins$bill_depth_mm)
```

Analyse the penguins dataset to investigate the hypothesis that flipper length is influenced by bill length and bill depth (at this point without considering interactions). Do so following a series of steps in the Bayesian workflow.

## Exercise 1

Translate the hypothesis into a Bayesian mathematical model. What is the general structure of this model?

Answer: Suppose flipper length is $y_{i}$, bill length is $x_{1i}$ and bill depth is $x_{2i}$.

$Y_{i} \sim Normal(\mu_{i},\ \sigma ^2),\ \mu_{i} = \beta_{0} + \beta_{1}x_{1i} + \beta_{2}x_{2i}$.

We need priors on all $\beta$ and $\sigma$.

## Exercise 2

Take a look at the default priors for the model, and define two sets of alternative priors.

```{r}
default_prior(flipper_length_mm ~ 1 + bill_length_mm + bill_depth_mm,
              data = penguins)

prior1 <- c(prior(normal(200, 20), class = Intercept), 
            prior(normal(0, 4), class = b, coef = bill_depth_mm),
            prior(normal(0, 5), class = b, coef = bill_length_mm),
            prior(normal(5, 2), class = sigma, lb = 0))

prior2 <- c(prior(normal(100, 20), class = Intercept), 
            prior(normal(0, 1), class = b, coef = bill_depth_mm),
            prior(normal(0, 1), class = b, coef = bill_length_mm),
            prior(normal(5, 5), class = sigma, lb = 0))
```

## Exercise 3

Conduct a small sensitivity analysis by comparing the prior predictive values of the model for the three sets of priors.

```{r}
#| eval: false

# default prior
file_path <- "brms_prior1.rds"
if (file.exists(file_path)) {
  model.prior1 <- readRDS(file_path)
} else {
  set.seed(359)
  model.prior1 <- brm(flipper_length_mm ~ 
                        1 + bill_length_mm + bill_depth_mm,
                      data = penguins, 
                      sample_prior = "only", iter = 2000, cores = 4)
  saveRDS(model.prior1, file = file_path)
}
prior.predictions <- posterior_predict(model.prior1)
bayesplot::ppc_dens_overlay(prior.predictions[1,], 
                            prior.predictions[1:500,], trim = T) +
  theme_light(base_size = 16)
```

No prior predictive values can be obtained when the priors are flat, because flat priors are "improper", i.e. they violate the axioms of probability theory. This is a reason for choosing proper priors instead.

```{r}
# defined set 1
file_path <- "brms_prior2.rds"
if (file.exists(file_path)) {
  model.prior2 <- readRDS(file_path)
} else {
  set.seed(359)
  model.prior2 <- brm(flipper_length_mm ~ 
                        1 + bill_length_mm + bill_depth_mm,
                      data = penguins, prior = prior1,
                      sample_prior = "only", iter = 2000, cores = 4)
  saveRDS(model.prior2, file = file_path)
}
pp_check(model.prior2)

# defined set 2
file_path <- "brms_prior3.rds"
if (file.exists(file_path)) {
  model.prior3 <- readRDS(file_path)
} else {
  set.seed(359)
  model.prior3 <- brm(flipper_length_mm ~ 
                        1 + bill_length_mm + bill_depth_mm,
                      data = penguins, prior = prior2,
                      sample_prior = "only", iter = 2000, cores = 4)
  saveRDS(model.prior3, file = file_path)
}
pp_check(model.prior3)
```

Prior1 is a more reasonable set.

## Exercise 4

Estimate the model using the priors that was better in the prior predictive checks. Assess the convergence parameters and briefly interpret the model output.

```{r}
file_path <- "brms_model1.rds"
if (file.exists(file_path)) {
  model1 <- readRDS(file_path)
} else {
  set.seed(359)
  model1 <- brm(flipper_length_mm ~ 1 + bill_length_mm + bill_depth_mm,
                data = penguins, prior = prior1,
                iter = 2000, warmup = 500, silent = 2, refresh = 0)
  saveRDS(model1, file = file_path)
}
model1
plot(model1)
```

Model diagnostics: The effective sample sizes for each parameter are above 1000, and the Gelmann-Rubin criterion R_hat is not larger than 1. The four chains have mixed well, showing a horizontal zig-zag with no discernable systematic drift upwards or downwards. The posterior distributions for each parameter are unimodal and relatively smooth; and they occupy limited regions of the parameter space rather than stretching across all or most possible parameter values. All this speaks in favour of the model having converged to a stable solution.

Interpretation: None of the credible intervals include the Null. Bill length seems to have a slightly positive effect, and bill depth a slightly negative effect on flipper length.

## Exercise 5

Define two alternative models: An intercept-only model without predictors, and a more complex model that includes not just the main effects but also their interaction. Compare these models using Bayes factors, and briefly interpret the results.

```{r}
# null model
prior_null <- c(prior(normal(200, 20), class = Intercept),
            prior(normal(5, 2), class = sigma, lb = 0))

file_path <- "brms_model0.rds"
if (file.exists(file_path)) {
  model0 <- readRDS(file_path)
} else {
  set.seed(359)
  model0 <- brm(flipper_length_mm ~ 1,
                data = penguins, prior = prior_null,
                iter = 2000, warmup = 500, silent = 2, refresh = 0)
  saveRDS(model0, file = file_path)
}
model0
plot(model0)

# full model with interaction
prior_full <- c(prior(normal(200, 20), class = Intercept), 
            prior(normal(0, 4), class = b, coef = bill_depth_mm),
            prior(normal(0, 5), class = b, coef = bill_length_mm),
            prior(normal(0, 5), class = b, coef = bill_length_mm:bill_depth_mm),
            prior(normal(5, 2), class = sigma, lb = 0))

file_path <- "brms_model2.rds"
if (file.exists(file_path)) {
  model2 <- readRDS(file_path)
} else {
  set.seed(359)
  model2 <- brm(flipper_length_mm ~ 1 + bill_length_mm * bill_depth_mm,
                data = penguins, prior = prior_full,
                iter = 2000, warmup = 500, silent = 2, refresh = 0)
  saveRDS(model2, file = file_path)
}
model2
plot(model2)

# model comparison
bayes_factor(model1, model0)
bayes_factor(model2, model1)
```

Interpretation: There is very strong evidence (BF \> 100) that model2 can account for the pattern in the data better than model1, which in turn can account for the data better than model0.

## Exercise 6

Conduct posterior predictive checks for the model that fairs best on the basis of the Bayes factor comparison.

```{r}
posterior.predictions <- posterior_predict(model2)
bayesplot::ppc_dens_overlay(penguins$flipper_length_mm,
                            posterior.predictions[1:500,]) + 
  theme_light(base_size = 16)

# alternatively
pp_check(model2)
```
