---
title: "Assignment7"
format: html
editor: visual
---

```{r setup}
#| message: false
#| warning: false
  
# packages
library(tidyverse) # Graphs with ggplot()
library(knitr) # Tables with kable()
library(brms) # Hierarchical Bayesian models
library(rstan) # Interface to Stan
library(bayesplot) # Model output graphs

set.seed(359)
```

**Are subject relatives easier to process than object relatives (shifted log-normal likelihood)?**

This is a classic question from the psycholinguistics literature: Are subject relatives easier to process than object relatives? The data come from Experiment 1 in a paper by Grodner and Gibson 2005.

*Scientific question: Is there a subject relative advantage in reading?*

Grodner and Gibson 2005 investigate an old claim in psycholinguistics that object relative clause (ORC) sentences are more difficult to process than subject relative clause (SRC) sentences. One explanation for this predicted difference is that the distance between the relative clause verb (sent in the example below) and the head noun phrase of the relative clause (reporter in the example below) is longer in ORC vs. SRC.

Examples are shown below. The relative clause is shown in square brackets.

(1a) The reporter \[who the photographer sent to the editor\] was hoping for a good story. (ORC)

(1b) The reporter \[who sent the photographer to the editor\] was hoping for a good story. (SRC)

The underlying explanation has to do with memory processes: Shorter linguistic dependencies are easier to process due to either reduced interference or decay, or both.

In the Grodner and Gibson data, the dependent measure is reading time at the relative clause verb, (e.g., sent) of different sentences with either ORC or SRC. The dependent variable is in milliseconds and was measured in a self-paced reading task. Self-paced reading is a task where subjects read a sentence or a short text word-by-word or phrase-by-phrase, pressing a button to get each word or phrase displayed; the preceding word disappears every time the button is pressed.

For this experiment, we are expecting longer reading times at the relative clause verbs of ORC sentences in comparison to the relative clause verb of SRC sentences.

```{r}
#| message: false
#| warning: false

load("df_gg05_rc.rda")
```

You should use a sum coding for the predictors. Here, object relative clauses ("objgaps") are coded +1/2, subject relative clauses −1/2.

```{r}
#| message: false
#| warning: false

df_gg05_rc <- df_gg05_rc %>%
  mutate(c_cond = if_else(condition == "objgap", 1/2, -1/2))
```

Before fitting, also clean the reading times: exclude relative clause verb reading times below 150 ms, since responses that fast probably are an anticipatory button-press instead of actually reading the target word and reacting. Report how many trials this excludes.

```{r}
#| message: false
#| warning: false

n_before <- nrow(df_gg05_rc)
df_gg05_rc <- df_gg05_rc %>% 
  filter(RT >= 150)
n_before - nrow(df_gg05_rc)
```

Reading times can never be shorter than some minimum time (the time needed for basic perceptual other processes necessary for reading). We can capture this by shifting the log-normal distribution away from zero by an additional parameter (ndt, non-decision time). You can now fit a “maximal” model (correlated varying intercept and slopes for subjects and for items) assuming a shifted log-normal likelihood.

## Exercise 1

\(a\) Specify the statistical model and define priors for all parameters in the model, including the additional shift parameter ndt.

Hint: To get a feel for reasonable ranges of values for the µ and σ parameters of a (shifted) log-normal distribution, you might want to simulate log-normal data for different values using `rlnorm()` and then add a constant shift to it. Also think about what constraints ndt needs to satisfy relative to the observed data for the model to be identifiable, and where the bound for its prior should come from. Should it be extracted from this particular sample, or fixed in advance?

**Caution:** brms uses a log link for ndt by default (check `?brms::shifted_lognormal`), which means that, depending on how you specify the model, the units your prior and posterior for ndt are expressed in may not be milliseconds. Before interpreting or setting a prior for ndt, check `make_stancode()` for your model to see exactly how ndt is parametrized.

#### Do I need random effect at all?

```{r}
#| message: false
#| warning: false

# individual
g1 <- ggplot(df_gg05_rc, aes(x = c_cond, y = RT, 
                 color = factor(subj), 
                 fill = factor(subj)))+
  geom_point()+
  geom_smooth(method = "lm") +
  theme_bw()+
  labs(x = "Condition", y = "Reading time", 
       color = "Subject ID", fill = "Subject ID")
g1

# item
g2 <- ggplot(df_gg05_rc, aes(x = c_cond, y = RT, 
                             color = factor(item), 
                             fill = factor(item)))+
  geom_point()+
  geom_smooth(method = "lm") +
  theme_bw()+
  labs(x = "Condition", y = "Reading time", 
       color = "Item ID", fill = "Item ID")
g2
```

There seems no sufficient reason to use item-level random effects.

#### Define model

```{r}
#| message: false
#| warning: false

default_prior(RT ~ c_cond + (c_cond | subj) + (c_cond | item), 
              family = shifted_lognormal(), 
              data = df_gg05_rc)

prior1 <- c(prior(normal(6, 1), class = Intercept), 
            # mean of log reading time
            prior(normal(0, 0.1), class = b, coef = c_cond), 
            prior(normal(0, 1), class = sigma, lb = 0), 
            prior(normal(0, 1), class = sd),
            prior(lkj(2), class = cor),
            prior(uniform(0, 150), class = ndt, lb = 0, ub = 150))

# full model
file_path <- "brms_model1.rds"
if (file.exists(file_path)) {
  model1 <- readRDS(file_path)
} else {
  model1 <- brm(RT ~ c_cond + (c_cond | subj) + (c_cond | item),
                family = shifted_lognormal(), 
                data = df_gg05_rc, prior = prior1,
                iter = 2000, warmup = 500, cores = 4,
                save_pars = save_pars(all = TRUE))
  saveRDS(model1, file = file_path)
}
model1
```

Calculate ICC:

Between subject variance: $\sigma^2_{subj} = 0.39^2 = 0.1521$.

Between item variance: $\sigma^2_{item} = 0.05^2 = 0.0025$.

Residuals: $\sigma^2_{e} = 0.59^2 = 0.3481$.

ICC of subjects: $ICC\_{subj} = 0.1521 / (0.1521 + 0.0025 + 0.3481) \approx 0.30$.

ICC of items: $ICC\_{item} = 0.0025 / (0.1521 + 0.0025 + 0.3481) \approx 0.005$.

Again, there seems no reason to include item-level random effects.

## Exercise 2

Examine the effect of relative clause attachment site (the predictor c_cond) on reading times RT (β1).

Answer: Relative clause has no clear effect on reading times, 95% CI: \[-0.04, 0.23\] but they vary a lot between subjects.

## Exercise 3

Estimate the median difference between relative clause attachment sites in milliseconds, and report the mean and 95% CI. Remember to take the estimated shift $ndt$ into account when you transform back to the millisecond scale.

Hint: For a shifted log-normal, the expected reading time in condition X is $ndt + exp(\beta_{0} + \beta_{1}X)$. Because $ndt$ is a constant additive shift that is the same for both conditions, it cancels out when we compute the difference between conditions, so the difference formula is the same as if we would have a log-normal model.

```{r}
#| message: false
#| warning: false

draws <- as_draws_df(model1)
beta0 <- draws$b_Intercept
beta1 <- draws$b_c_cond
ndt <- draws$ndt

effect <- exp(beta0 + beta1 * .5) - exp(beta0 + beta1 * -.5)
c(mean = mean(effect), quantile(effect, c(.025, .975)))

rt_objgap <- ndt + exp(beta0 + beta1 * .5)
rt_subjgap <- ndt + exp(beta0 + beta1 * -.5)
c(mean_objgap = mean(rt_objgap), mean_subjgap = mean(rt_subjgap))
```

Answer: The estimated reading time difference between the two conditions is 17.80ms, 95% CI: \[-6.00, 43.59\]. The mean reading time in the objective group is 337.46ms, and 319.66ms in the subjective group.

#### Consider no item-level random effect

```{r}
#| message: false
#| warning: false

file_path <- "brms_model0.rds"
if (file.exists(file_path)) {
  model0 <- readRDS(file_path)
} else {
  model0 <- brm(RT ~ c_cond + (c_cond | subj),
                family = shifted_lognormal(), 
                data = df_gg05_rc, prior = prior1,
                iter = 2000, warmup = 500, cores = 4,
                save_pars = save_pars(all = TRUE))
  saveRDS(model0, file = file_path)
}
model0

bayes_factor(model0, model1)
```

The model without item-level random effects actually performs better.

## Exercise 4

Inspect individual differences in the subject relative advantage effect. Are there participants with effects in opposite directions?

```{r}
#| message: false
#| warning: false

by_subj_effect <- data.frame(coef(model1)$subj[,, "c_cond"]) %>%
  bind_rows() %>%
  mutate(subj = unique(df_gg05_rc$subj)) %>%
  arrange(Estimate) %>%
  mutate(subj = factor(subj, levels = subj))

ggplot(
  by_subj_effect,
  aes(ymin = Q2.5, ymax = Q97.5, x = subj, y = Estimate)
  ) +
  geom_errorbar() +
  geom_point() +
  geom_hline(
    yintercept = posterior_summary(model1)["b_c_cond", "Estimate"]
    ) +
  geom_hline(
    yintercept = posterior_summary(model1)["b_c_cond", "Q2.5"],
    linetype = "dotted", linewidth = 0.5
    ) +
  geom_hline(
    yintercept = posterior_summary(model1)["b_c_cond", "Q97.5"],
    linetype = "dotted", linewidth = 0.5
    ) +
  xlab("Change in reading times") +
  coord_flip() +
  theme_light(base_size = 16)
```

As before, some subjects' credible intervals cross zero or point in the direction opposite to the group-level estimate, showing that not everyone shows the same (or even same-signed) relative advantage.

We may also look at per-item difference.

```{r}
#| message: false
#| warning: false

by_item_effect <- data.frame(coef(model1)$item[,, "c_cond"]) %>%
  bind_rows() %>%
  mutate(item = unique(df_gg05_rc$item)) %>%
  arrange(Estimate) %>%
  mutate(item = factor(item, levels = item))

ggplot(
  by_item_effect,
  aes(ymin = Q2.5, ymax = Q97.5, x = item, y = Estimate)
  ) +
  geom_errorbar() +
  geom_point() +
  geom_hline(
    yintercept = posterior_summary(model1)["b_c_cond", "Estimate"]
    ) +
  geom_hline(
    yintercept = posterior_summary(model1)["b_c_cond", "Q2.5"],
    linetype = "dotted", linewidth = 0.5
    ) +
  geom_hline(
    yintercept = posterior_summary(model1)["b_c_cond", "Q97.5"],
    linetype = "dotted", linewidth = 0.5
    ) +
  xlab("Change in reading times") +
  coord_flip() +
  theme_light(base_size = 16)
```

## Exercise 5

Do a sensitivity analysis. What is the estimate of the effect (β1) under different priors? What is the difference in ms between conditions under different priors? Does the choice of prior on $ndt$ also affect the results?

Note: In the sensitivity analysis, model0 is used as the benchmark.

#### Coefficient (β1)

```{r}
#| message: false
#| warning: false

# model2
prior2 <- c(prior(normal(6, 1), class = Intercept), 
            # mean of log reading time
            prior(normal(0, 1), class = b, coef = c_cond), 
            prior(normal(0, 1), class = sigma, lb = 0), 
            prior(normal(0, 1), class = sd),
            prior(lkj(2), class = cor),
            prior(uniform(0, 150), class = ndt, lb = 0, ub = 150))

file_path <- "brms_model2.rds"
if (file.exists(file_path)) {
  model2 <- readRDS(file_path)
} else {
  model2 <- brm(RT ~ c_cond + (c_cond | subj),
                family = shifted_lognormal(), 
                data = df_gg05_rc, prior = prior2,
                iter = 2000, warmup = 500, cores = 4,
                save_pars = save_pars(all = TRUE))
  saveRDS(model2, file = file_path)
}
model2

# model3
prior3 <- c(prior(normal(6, 1), class = Intercept), 
            # mean of log reading time
            prior(normal(0, 0.01), class = b, coef = c_cond), 
            prior(normal(0, 1), class = sigma, lb = 0), 
            prior(normal(0, 1), class = sd),
            prior(lkj(2), class = cor),
            prior(uniform(0, 150), class = ndt, lb = 0, ub = 150))

file_path <- "brms_model3.rds"
if (file.exists(file_path)) {
  model3 <- readRDS(file_path)
} else {
  model3 <- brm(RT ~ c_cond + (c_cond | subj),
                family = shifted_lognormal(), 
                data = df_gg05_rc, prior = prior3,
                iter = 2000, warmup = 500, cores = 4,
                save_pars = save_pars(all = TRUE))
  saveRDS(model3, file = file_path)
}
model3
```

Comparing $\beta_1$ across priors:

```{r}
list(model3, model0, model2) %>%
  map_dfr(~
  posterior_summary(.x, variable = "b_c_cond") %>%
    as_tibble() %>%
    select(-Est.Error)) %>%
  mutate("Prior for $\\beta_1$" = c("Normal(0,.001)", 
                                    "Normal(0,0.1)", "Normal(0,1)")) %>%
  knitr::kable(escape = FALSE)
```

And the corresponding difference in milliseconds (`ndt` cancels out):

```{r}
list(model3, model0, model2) %>%
  map_dfr(~ {
    d <- as_draws_df(.x)
    diff <- exp(d$b_Intercept + d$b_c_cond / 2) -
      exp(d$b_Intercept - d$b_c_cond / 2)
    c(mean = mean(diff), quantile(diff, c(.025, .975))) %>%
      bind_rows() %>%
      setNames(c("Estimate (ms)", "Q2.5", "Q97.5"))
  }) %>%
  mutate("Prior for $\\beta_1$" = c("Normal(0,0.01)", 
                                    "Normal(0,0.1)", "Normal(0,1)")) %>%
  knitr::kable(escape = FALSE)
```

As in a log-normal model, the posterior for $\beta_{1}$ and the (implied) millisecond difference change a lot depending on how informative the prior is, especially when the effect is small relative to the data.

#### ndt

```{r}
#| message: false
#| warning: false

# model4
prior4 <- c(prior(normal(6, 1), class = Intercept), 
            # mean of log reading time
            prior(normal(0, 0.1), class = b, coef = c_cond), 
            prior(normal(0, 1), class = sigma, lb = 0), 
            prior(normal(0, 1), class = sd),
            prior(lkj(2), class = cor),
            prior(normal(50, 50), class = ndt, lb = 0, ub = 150))

file_path <- "brms_model4.rds"
if (file.exists(file_path)) {
  model4 <- readRDS(file_path)
} else {
  model4 <- brm(RT ~ c_cond + (c_cond | subj),
                family = shifted_lognormal(), 
                data = df_gg05_rc, prior = prior4,
                iter = 2000, warmup = 500, cores = 4,
                save_pars = save_pars(all = TRUE))
  saveRDS(model4, file = file_path)
}
model4

list(model0, model4) %>%
  map_dfr(~ {
    d <- as_draws_df(.x)
    diff <- exp(d$b_Intercept + d$b_c_cond / 2) -
      exp(d$b_Intercept - d$b_c_cond / 2)
    tibble(
      "ndt Estimate" = mean(d$ndt),
      "beta1 Estimate" = mean(d$b_c_cond),
      "Diff (ms)" = mean(diff)
    )
  }) %>%
  mutate("Prior for ndt" = c("Uniform(0, 150ms)", "Normal+(50, 50)")) %>%
  knitr::kable(escape = FALSE)
```

Because `ndt` cancels out of the difference between conditions, changing its prior mostly shifts the estimate of `ndt` (and also $\beta_0$/$\sigma$), but has little effect on $\beta_1$ or on the millisecond difference between conditions.

## Exercise 6

So far ndt has been a single constant shared by everyone. In reality, people probably differ in their non-decision time too. Extend the model so that $ndt$ varies by subject `(ndt ~ 1 + (1 | subj))`, and think carefully about what prior to put on it. Keep in mind the log link caution from part (a).

Hint: a prior with mean around 5.5 to 6.5 on the linear predictor scale for $ndt$’s intercept, together with a weakly-informative prior on its between-subject SD, could work out well here. Think about why, and what those numbers correspond to once exponentiated back to milliseconds. You also need to to give the sampler some sensible starting values.

```{r}
#| message: false
#| warning: false
#| eval: false

bprior_ndt <- c(
  prior(normal(5, 0.7), class = Intercept),
  prior(normal(0, .1), class = b),
  prior(normal(0, 1), class = sigma),
  prior(normal(0, 1), class = sd),
  prior(lkj(2), class = cor),
  prior(normal(6, 0.5), class = Intercept, dpar = ndt, lb = 0),
  prior(exponential(1), class = sd, dpar = ndt)
)

fit_shifted_indiv_ndt <- brm(
  bf(RT ~ c_cond + (c_cond | subj) + (c_cond | item),
     ndt ~ 1 + (1 | subj)),
  family = shifted_lognormal(),
  prior = bprior_ndt,
  data = df_gg05_rc,
  init = 0.1,
  control = list(adapt_delta = 0.99, max_treedepth = 15)
)
```

- `normal(6, 0.5)` on `ndt`'s intercept looks like it implies a very tiny non-decision time, but it doesn't because of the log link. `6` is on the linear predictor scale, and the corresponding non-decision time is $\exp(6) \approx 403$ ms, so a plausible value for reading. The SD of $0.5$ means the bulk of the prior mass for an individual subject falls somewhere between $\exp(5)\approx 148$ and $\exp(7)\approx 1097$ ms.
- `lb = 0` bounds the linear predictor, not `ndt`. Since `ndt` is already positive after exponentiating, this barely has an effect, it only rules out non-decision times below $\exp(0)=1$ ms. It is not the same kind of hard bound we used for the fixed-`ndt` model.
- The by-subject SD gets a weakly informative `exponential(1)` prior, this is a common prior for a group-level SD on a log scale.
- Here is no simple way to give each subject's `ndt` a per-subject upper bound tied to their own minimum RT. Identifiability relies on the prior being concentrated well below the observed RTs, so implausible regions get basically no prior mass.
- A badly-chosen starting point can effect the sampler to sample in an invalid region (where $Y_{ik} - \mathit{ndt}_i < 0$ for some subject), so it helps to set starting values, e.g. `init = 0.1`, instead of leaving `init` at its default.